---
title: "NC births"
author: "Amelia McNamara"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.show = "hold", out.width = "50%")
library(tidyverse)
library(openintro)
```



An attempt to reproduce the `ncbirths` dataset from the `openintro` package. Data provenance has been lost on that dataset, other than the fact that it's from 2004. 

I found the [ICPSR data from 2004](https://www.icpsr.umich.edu/web/ICPSR/studies/4707/versions/V1), which is quite large. 
```{r data-load, eval=FALSE}
X04707_0001_Data <- read_delim("ICPSR_04707/DS0001/04707-0001-Data.tsv",
  "\t", escape_double = FALSE, trim_ws = TRUE)
nrow(X04707_0001_Data)
```

Not going to run that code when I knit because it's so big. `nrow` should be 4118907. 

## North Carolina

Then, I filtered down to just the babies from North Carolina. 
```{r nc, eval=FALSE}
nc <- filter(X04707_0001_Data, OSTATE == "NC")
nrow(nc)
```

For the record, `nrow` there should be 120588. 

Let's write out just those babies for the future,

```{r write-nc, eval=FALSE}
write_csv(nc, "nc_full.csv")
```

And, for our purposes in the document, we can start there,

```{r}
nc <- read_csv("nc_full.csv")
```

## Cleaning 

Using the data dictionary, I renamed the variables that looked relevant, and selected those out.  
```{r rename}
nc <- nc %>%
  rename(
    fage = FAGEREC11,
    mage = MAGER14,
    weeks = COMBGEST,
    visits = UPREVIS,
    gained = WTGAIN,
    weight_g = DBWT,
    sex = SEX,
    habit = CIG_REC6,
    marital = MAR,
    whitemom = MRACEREC
  ) %>%
  select(fage, mage, weeks, visits, 
         gained, weight_g, sex, habit, marital, whitemom)
```

Again using the data dictionary, I recoded most of the variables. 
```{r}
nc <- nc %>%
  mutate(
    fage_cat = case_when(
      fage == 1 ~ "Under 15 years",
      fage == 2 ~ "15-19 years",
      fage == 3 ~ "20-24 years",
      fage == 4 ~ "25-29 years",
      fage == 5 ~ "30-34 years",
      fage == 6 ~ "35-39 years",
      fage == 7 ~ "40-44 years",
      fage == 8 ~ "45-49 years",
      fage == 9 ~ "50-54 years",
      fage == 10 ~ "55-98 years",
      fage == 11 ~ NA_character_
    ),
    mage_cat = case_when(
      mage == 1 ~ "Under 15 years",
      mage == 3 ~ "15",
      mage == 4 ~ "16",
      mage == 5 ~ "17",
      mage == 6 ~ "18",
      mage == 7 ~ "19",
      mage == 8 ~ "20-24 years",
      mage == 9 ~ "25-29 years",
      mage == 10 ~ "30-34 years",
      mage == 11 ~ "35-39 years",
      mage == 12 ~ "40-44 years",
      mage == 13 ~ "45-49 years",
      mage == 14 ~ "50-54 years"
    ),
    mature = case_when(
      mage %in% c("35-39 years", "40-44 years", "45-49 years", "50-54 years") ~ "mature mom",
      TRUE ~ "younger mom"
    ),
    weeks = case_when(
      weeks == 99 ~ NA_real_,
      TRUE ~ weeks
    ),
    # could read 99 in as NA in read_delim?
    premie = case_when(
      weeks < 37 ~ "premie",
      weeks >= 37 ~ "full term",
      is.na(weeks) ~ NA_character_
    ),
    visits = case_when(
      visits == 99 ~ NA_real_,
      TRUE ~ visits
    ),
    gained = case_when(
      gained == 99 ~ NA_real_,
      TRUE ~ gained
    ),
    weight = weight_g / 453.59237,
    lowbirthweight = case_when(
      weight <= 5.5 ~ "low",
      TRUE ~ "not low"
    ),
    sex = case_when(
      sex == 1 ~ "male",
      sex == 2 ~ "female"
    ),
    habit = case_when(
      habit == 0 ~ "nonsmoker",
      habit == 6 ~ NA_character_,
      TRUE ~ "smoker"
    ),
    marital = case_when(
      marital == 1 ~ "married",
      marital == 2 ~ "not married",
      marital == 9 ~ NA_character_
    ),
    whitemom = case_when(
      whitemom == 1 ~ "white",
      TRUE ~ "not white"
    )
  )
nc <- nc %>%
  select(fage_cat, mage_cat, mature, weeks, premie, visits, 
         gained, weight, lowbirthweight, sex, habit, 
         marital, whitemom)
```

We want to simulate ages from the categorical variables `mage_cat` and `fage_cat`, so we need a little function: 
```{r simulate-age}
simulate_ages <- function(.x, .y) {
  if(is.na(.y)) {
    rep(NA_integer_, times = nrow(.x))
  } else if (.y == "Under 15 years") {
    sample(11:14, size = nrow(.x), replace = TRUE)
  } else if (.y %in% c("15", "16", "17", "18", "19")) {
    rep(as.numeric(.y), times = nrow(.x))
  } else if (.y == "55-98 years") {
    sample(55:98, size = nrow(.x), replace = TRUE)
  } else {
    age <- as.numeric(substr(.y, start = 1, stop = 2))
    sample(age:(age + 4), size = nrow(.x), replace = TRUE)
  }
}
```

Use the function: 
```{r}
nc <- nc %>%
  group_by(mage_cat) %>%
  nest() %>%
  mutate(mage = map2(data, mage_cat, simulate_ages)) %>%
  unnest(cols = c(data, mage)) %>%
  ungroup()
nc <- nc %>%
  group_by(fage_cat) %>%
  nest() %>%
  mutate(fage = map2(data, fage_cat, simulate_ages)) %>%
  unnest(cols = c(data, fage)) %>%
  ungroup()



dim(nc)
```

## Random sample
Because `ncbirths` is only 1,000 rows, I took a random sample. 

```{r}
set.seed(42)
nc_sample <- nc %>%
  sample_n(1000)
data(ncbirths)
```


## Comparing to ncbirths 
Then, I did all the univariate plots to compare my sample to `ncbirths`. 


For age, we now have two variables: a categorical one from the original data, and a numeric one we simulated. 

Let's start by comparing the original numeric variables to the categorical ones. 

Father's age
```{r}
ggplot(ncbirths) +
  geom_histogram(aes(fage))
ggplot(nc_sample) +
  geom_bar(aes(fage_cat))
```

Mother's age
```{r}
ggplot(ncbirths) +
  geom_histogram(aes(mage))
ggplot(nc_sample) +
  geom_bar(aes(mage_cat))
```

Now, let's look at our distributions with the simulated variables,

Father's age
```{r}
ggplot(ncbirths) +
  geom_histogram(aes(fage))
ggplot(nc_sample) +
  geom_histogram(aes(fage))
```

Mother's age
```{r}
ggplot(ncbirths) +
  geom_histogram(aes(mage))
ggplot(nc_sample) +
  geom_histogram(aes(mage))
```

These are still a bit weird. 

Beyond that, the distributions are pretty comparable 

```{r}
ggplot(ncbirths) +
  geom_bar(aes(x = mature))
ggplot(nc_sample) +
  geom_bar(aes(x = mature))
```

```{r}
ggplot(ncbirths) +
  geom_histogram(aes(weeks))
ggplot(nc_sample) +
  geom_histogram(aes(weeks))
```

We randomly didn't sample any NAs here, so that's a little different
```{r}
ggplot(ncbirths) +
  geom_bar(aes(x = premie))
ggplot(nc_sample) +
  geom_bar(aes(x = premie))
```

```{r}
ggplot(ncbirths) +
  geom_histogram(aes(visits))
ggplot(nc_sample) +
  geom_histogram(aes(visits))
```

```{r}
ggplot(ncbirths) +
  geom_histogram(aes(gained))
ggplot(nc_sample) +
  geom_histogram(aes(gained))
```

```{r}
ggplot(ncbirths) +
  geom_histogram(aes(weight))
ggplot(nc_sample) +
  geom_histogram(aes(weight))
```

```{r}
ggplot(ncbirths) +
  geom_bar(aes(x = lowbirthweight))
ggplot(nc_sample) +
  geom_bar(aes(x = lowbirthweight))
```

```{r}
ggplot(ncbirths) +
  geom_bar(aes(x = gender))
ggplot(nc_sample) +
  geom_bar(aes(x = sex))
```

```{r}
ggplot(ncbirths) +
  geom_bar(aes(x = habit))
ggplot(nc_sample) +
  geom_bar(aes(x = habit))
```

Again, we didn't randomly sample NAs for marital status
```{r}
ggplot(ncbirths) +
  geom_bar(aes(x = marital))
ggplot(nc_sample) +
  geom_bar(aes(x = marital))
```

or race
```{r}
ggplot(ncbirths) +
  geom_bar(aes(x = whitemom))
ggplot(nc_sample) +
  geom_bar(aes(x = whitemom))
```

## Other checks

Following [Mine's lab](http://www2.stat.duke.edu/~mc301/ARTSCI101_Su16/post/r/nc_births.html) I wanted to make sure things look similar enough there.

```{r}
ncbirths %>%
  summarise(mean_wt = mean(weight), sd_wt = sd(weight), n = n())
nc_sample %>%
  summarise(mean_wt = mean(weight), sd_wt = sd(weight), n = n())
```

Those look close to me! 

## Bivariate relationships

```{r}
ggplot(data = ncbirths, aes(x = habit, y = weight)) +
  geom_boxplot()
ggplot(data = nc_sample, aes(x = habit, y = weight)) +
  geom_boxplot()
```

Let's see how the ages look in relation to one another
```{r}
ggplot(ncbirths) + geom_point(aes(x=fage, y=mage))
ggplot(nc_sample) + geom_point(aes(x=fage, y=mage))
```

Okay, part of the problem is that outlier in father's age. 
```{r}
ggplot(ncbirths) + geom_point(aes(x=fage, y=mage)) + xlim(10, 60)
ggplot(nc_sample) + geom_point(aes(x=fage, y=mage)) + xlim(10, 60)
```

For a little more rigor, we could look at the linear model, 
```{r}
m1 <- lm(mage~fage, data = ncbirths)
m2 <- lm(mage~fage, data = filter(nc_sample, fage < 60))
confint(m1)
confint(m2)
```

Ooh, that's not great... the confidence intervals don't even intersect! 

```{r}
ncbirths %>%
  summarize(cor(fage, mage, use = "complete.obs"))
nc_sample %>%
  summarize(cor(fage, mage, use = "complete.obs"))
```

Maybe the simulation needs to take into account both father and mother at the same time. 


```{r}
ncbirths %>%
  group_by(habit) %>%
  summarise(mean_weight = mean(weight))
nc_sample %>%
  group_by(habit) %>%
  summarise(mean_weight = mean(weight))
```

Again, looks close.

## Writing data

Let's write out that small dataset,

```{r}
write_csv(nc_sample, "nc_sample.csv")
```
